---
title: "Analysis of the 2020-2021 Premier League Season"
output: html_document
---

<p>&nbsp;</p> <!--Adds vertical space -->


<div id = "top">
## Introduction
***************

<p>
The Premier League, often referred to outside the UK as the English Premier League, or sometimes the EPL, is the top level of the English football league system. Contested by 20 clubs, it operates on a system of promotion and relegation with the English Football League.
</p>

<p>
In this project, I have attempted to visualize a few of the most common football stats as seen on major sports channels and publications. The raw data used can be found on [this](http://www.football-data.co.uk/englandm.php) website. 
</p>
</div>

<p>&nbsp;</p>

<div id = "links">
## Quick Links
**************
* <a href ="#gnc" >Getting and Cleaning Data</a>
* <a href ="#stats" >Performing Statistical Tests</a>
* <a href ="#visualize" >Visualizing the Data</a>
  + <a href ="#points" >Points</a>
  + <a href ="#scored" >Goals Scored</a>
  + <a href ="#conceded" >Goals Conceded</a>
  + <a href ="#pvs" >Points Vs Goals Scored</a>
  + <a href ="#pvc" >Points Vs Goals Conceded</a>
  + <a href ="#home" >Home Performance</a>
  + <a href ="#away" >Away Performance</a>
  + <a href ="#creativity" >Creativity</a>
  + <a href ="#clinical" >Clinical Attacks</a>
  + <a href ="#fouls" >Fair Play/ Tactical Fouls</a>
  + <a href ="#ycards" >Yellow Cards</a>
  + <a href ="#rcards" >Red Cards</a>
</div>

<p>&nbsp;</p>
  
<div id = "gnc">
## Getting and Cleaning the Data
********************************

Fetching the data from the github repository

```{r, message = FALSE}
full.data <- read.csv("https://raw.githubusercontent.com/rohanprad/Data-Science-Project/main/Dataset/PL20-21.csv")
```

<p>&nbsp;</p>

Selecting only relevant observations

```{r, message = FALSE}
library(dplyr)

tidy.data <- select(full.data, Date:AR)
tidy.data <- select(tidy.data, -(Time))
tidy.data <- select(tidy.data, -(Referee))
```

<p>&nbsp;</p>

Getting the List of the 20 Team Names

```{r, message = FALSE}
clubs <- unique(tidy.data$HomeTeam)
```

<p>&nbsp;</p>

Then I defined a function called
[getTeamData()](https://github.com/rohanprad/Data-Science-Project/blob/main/TidyingUpTheData.R) 
which was used to summarize the results of all the matches and return a single row dataframe of 
relevant data columns for each team

```{r, include = FALSE}

getTeamData <- function(teamName){

      Home <- filter(tidy.data, HomeTeam == teamName)
      
      Away <- filter(tidy.data, AwayTeam == teamName)
      
      home.wins <- 0
      home.losses <- 0
      home.draws <- 0
      
      away.wins <- 0
      away.losses <- 0
      away.draws <- 0
      
      for(result in Home$FTR){
          if(result == "H"){
              home.wins <- home.wins + 1
          }else if(result == "A"){
              home.losses <- home.losses + 1
          }else{
              home.draws <- home.draws + 1
          }
      }
      
      
      for(result in Away$FTR){
        if(result == "A"){
          away.wins <- away.wins + 1
        }else if(result == "H"){
          away.losses <- away.losses + 1
        }else{
          away.draws <- away.draws + 1
        }
      }
      
      wins <- home.wins + away.wins
      draws <- home.draws + away.draws
      losses <- home.losses + away.losses
      
      played <- wins + draws + losses
      points <- (3*wins) + (draws)
      
      goals.for <- sum(Home$FTHG) + sum(Away$FTAG)
      goals.against <- sum(Home$FTAG) + sum(Away$FTHG)
      
      goal.difference <- goals.for - goals.against
      
      home.win.rate <- home.wins/length(Home$FTR) * 100
      home.win.rate <- round(home.win.rate)
      
      away.win.rate <- away.wins/length(Away$FTR) * 100
      away.win.rate <- round(away.win.rate)
      
      overall.win.rate <- wins/played * 100
      overall.win.rate <- round(overall.win.rate)
      
      goals.for.avg <- goals.for/played
      goals.for.avg <- round(goals.for.avg, digits = 1)
      
      goals.against.avg <- goals.against/played
      goals.against.avg <- round(goals.against.avg, digits = 1)
      
      shots <- sum(Home$HS) + sum(Away$AS)
      shots.target <- sum(Home$HST) + sum(Away$AST)
      
      corners <- sum(Home$HC) + sum(Away$AC)
      
      fouls <- sum(Home$HF) + sum(Away$AF)
      
      yellow.cards <- sum(Home$HY) + sum(Away$AY)
      red.cards <- sum(Home$HR) + sum(Away$AR)
      bookings <- yellow.cards + red.cards
      
      team.data <- data.frame(teamName, played, wins, draws, losses, 
                              goal.difference, points, goals.for, goals.against, 
                              goals.for.avg, goals.against.avg, home.wins, home.draws, home.losses, home.win.rate, 
                              away.wins, away.draws, away.losses, away.win.rate, overall.win.rate, 
                              shots, shots.target, corners, fouls, bookings, yellow.cards, red.cards)
      
      names(team.data) <- c("Club", "P", "W", "D", "L", "GD", "Pts", "GF", "GA", 
                            "GF.Avg", "GA.Avg", "HW", "HD", "HL", "HW.Rate", 
                            "AW", "AD", "AL", "AW.Rate", "W.Rate", "S", "ST",
                            "C", "F", "B", "YC", "RC")
      team.data

}
```

<p>&nbsp;</p>

Then I created an empty dataframe (the league table) to store the final results for all 20 teams

```{r, message = FALSE}
league.table <- data.frame(matrix(nrow = 0, ncol = 27))
colnames(league.table) <- c("Club", "P", "W", "D", "L", "GD", "Pts", "GF", "GA", 
                            "GF.Avg", "GA.Avg", "HW", "HD", "HL", "HW.Rate", 
                            "AW", "AD", "AL", "AW.Rate", "W.Rate", "S", "ST", 
                            "C", "F", "B", "YC", "RC")
for(club in clubs){
  row <- getTeamData(club)
  league.table <- rbind(league.table, row)
}

```

<p>&nbsp;</p>

Adding two new columns for each team defining the colour and the city

```{r, message = FALSE}
league.table <- arrange(league.table, Club)

teamColors <- c("#EF0107", "#95BFE5", "#0057B8", "#6C1D45", "#034694", "#A7A5A6",
                "#003399", "#000000", "#FFCD00", "#003090", "#9C1310", "#6CABDD", 
                "#DA291C", "#241F20", "#EE2737", "#D71920", "#132257", "#122F67", 
                "#7A263A", "#FDB913")

teamCities <- factor(c("London", "Birmingham", "Brighton & Hove", "Burnley", 
                       "London", "London", "Liverpool", "London", "Leeds", "Leceister",
                       "Liverpool", "Manchester", "Manchester", "Newcastle", "Sheffield",
                       "Southampton", "London", "West Bromwich", "London", "Wolverhampton"))

league.table$Col = teamColors
league.table$City = teamCities
```

<p>&nbsp;</p>

Arranging the league table by points > goal difference > goals for

```{r, message = FALSE}

league.table <- arrange(league.table, desc(Pts), desc(GD), desc(GF))

```

<p>&nbsp;</p>

Converting the clubs column into a factor

```{r, message = FALSE}

league.table <- mutate(league.table, Club = as.factor(league.table$Club))

```

<p>&nbsp;</p>

Creating a CSV file to store the final tidy data

```{r, message = FALSE}

write.csv(league.table, file = "./Dataset/TidyData.csv")

```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "stats">

## Performing Statistical Tests
*******************************

```{r, message = FALSE, warning = FALSE}
summary(league.table)

table(league.table$City)

table(league.table$Club)

table(league.table$Club, league.table$City)
```
<p>&nbsp;</p>

Chi Squared Test on Club and City  
Null Hypothesis: The variables are dependent.

```{r, message = FALSE, warning = FALSE}
summary(table(league.table$Club, league.table$City))
```
p-value > 0.05 : Fails to provide any evidence
<p>&nbsp;</p>


Quantile for number of goals scored with 5% probablity
```{r, message = FALSE, warning = FALSE}
quantile(league.table$GF, .05)
```
<p>&nbsp;</p>

Quantile for number of goals scored with 5% and 95% probablity
```{r, message = FALSE, warning = FALSE}
quantile(league.table$GF, c(.05, .95))
```

<p>&nbsp;</p>

Quantile for number of goals scored with an interval of 25% in the probabilties
```{r, message = FALSE, warning = FALSE}
quantile(league.table$GF)
```

<p>&nbsp;</p>

Using the t.test and asking if the mean of points can be 40
```{r, message = FALSE, warning = FALSE}
t.test(league.table$Pts, mu = 40)
```
p-value <  0.05 so it's unlikely that mean is 40.  
Null hypothesis is rejected.

<p>&nbsp;</p>

Using the t.test and asking if the mean of points can be 40 with a confidence level of 99%
```{r, message = FALSE, warning = FALSE}
t.test(league.table$Pts, conf.level = 0.99, mu = 40)
```
p-value <  0.05 so it's unlikely that mean is 40.  
Null hypothesis is rejected.

<p>&nbsp;</p>

Calculating the confidence interval for the median of points using the Wilcox Test
```{r, message = FALSE, warning = FALSE}
wilcox.test(league.table$Pts, conf.int = TRUE)
```
<p>&nbsp;</p>

Using the Shapiro Test for normality
```{r, message = FALSE, warning = FALSE}
shapiro.test(league.table$Pts)
```
p-value > 0.05 so it's normally distributed.

<p>&nbsp;</p>

Checking if the correlation between points and the number of wins is significant
```{r, message = FALSE, warning = FALSE}
cor.test(league.table$Pts, league.table$W)
```
p-value < 0.05 so a significant correlation exists.

<p>&nbsp;</p>

Checking if the correlation between points and the number of yellow cards is significant
```{r, message = FALSE, warning = FALSE}
cor.test(league.table$Pts, league.table$YC)
```
p-value > 0.05 so a significant correlation does not exist.

<p>&nbsp;</p>

</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<div id = "visualize">

## Visualizing the Data
***********************

Loading the necessary packages

```{r,  message = FALSE}

library(ggplot2)

#install.packages("devtools")
#devtools::install_github("slowkow/ggrepel")

library(ggrepel)
```
<p>&nbsp;</p>

<div id = "points">
### Who has how many points?
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(Club, Pts))+
  geom_bar(stat = "identity",
           fill = league.table$Col)+
  theme_bw()+
  coord_flip()+
  labs(x = "Club",
       y = "Points",
       title = "Premier League 2020-21 Points")
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "scored">
### Goals Scored
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(Club, GF))+
  geom_bar(stat = "identity",
           fill = league.table$Col)+
  theme_bw()+
  coord_flip()+
  labs(x = "Club",
       y = "Goals",
       title = "Premier League 2020-21 Goals Scored")
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "conceded">

### Goals Conceded
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(Club, GA))+
  geom_bar(stat = "identity",
           fill = league.table$Col)+
  theme_bw()+
  coord_flip()+
  labs(x = "Club",
       y = "Goals",
       title = "Premier League 2020-21 Goals Conceded")
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "pvs">
### Points Vs Goals Scored
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(GF, Pts) )+
  geom_point(size = 4, col = league.table$Col)+
  geom_smooth(method = lm, se = F, formula = y ~ x)+
  theme_bw()+
  labs(
    x = "Goals",
    y = "Points",
    title = "Points Vs Goals Scored")+
  geom_text_repel(
    label= league.table$Club, 
    nudge_x = 0, nudge_y = -1.0, 
    aes(size = 2.5)
  )+
  scale_size_identity()
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "pvc">
### Points Vs Goals Conceded
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(GA, Pts))+
  geom_point(size = 4, col = league.table$Col)+
  geom_smooth(method = lm, se = F, formula = y ~ x)+
  theme_bw()+
  labs(
    x = "Goals",
    y = "Points",
    title = "Points Vs Goals Conceded")+
  geom_text_repel(
    label= league.table$Club, 
    nudge_x = 0, nudge_y = -1.0, 
    aes(size = 3)
  )+
  scale_size_identity()
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "home">
### Home Performance
****************************

```{r,  message = FALSE}
ggplot(league.table, aes(HW.Rate, Pts))+
  geom_point(size = 4, col = league.table$Col)+
  geom_smooth(method = lm, se = FALSE, formula = y ~ x)+
  theme_bw()+
  labs(
    x = "Win %",
    y = "Points",
    title = "Home Performance")+
  geom_text_repel(
    label= league.table$Club, 
    nudge_x = 0, nudge_y = -0.25, 
    aes(size = 3)
  )+
  scale_size_identity()
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "away">
### Away Performance
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(AW.Rate, Pts))+
  geom_point(size = 4, col = league.table$Col)+
  geom_smooth(method = lm, se = FALSE, formula = y ~ x)+
  theme_bw()+
  labs(
    x = "Win %",
    y = "Points",
    title = "Away Performance")+
  geom_text_repel(
    label= league.table$Club, 
    nudge_x = 0, nudge_y = -0.25, 
    aes(size = 3)
  )+
  scale_size_identity()
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "creativity">
### Creativity 
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(S, ST))+
  geom_point(size = 4, col = league.table$Col)+
  geom_smooth(method = lm, se = FALSE, formula = y ~ x)+
  theme_bw()+
  labs(
    x = "Shots",
    y = "Shots on Target",
    title = "Creativity")+
  geom_text_repel(
    label= league.table$Club, 
    nudge_x = 0, nudge_y = -0.25, 
    aes(size = 3)
  )+
  scale_size_identity()
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "clinical">
### Clinical Attacks
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(S, (GF/S * 100)))+
  geom_point(size = 4, col = league.table$Col)+
  geom_smooth(method = lm, se = FALSE, formula = y ~ x)+
  theme_bw()+
  labs(
    x = "Shots",
    y = "Score Rate",
    title = "Clinical Attacks")+
  geom_text_repel(
    label= league.table$Club, 
    nudge_x = 0, nudge_y = -0.25, 
    aes(size = 3)
  )+
  scale_size_identity()
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "fouls">
### Fair Play / Tactical Fouls 
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(F, B))+
  geom_point(size = 4, col = league.table$Col)+
  geom_smooth(method = lm, se = FALSE, formula = y ~ x)+
  theme_bw()+
  labs(
    x = "Fouls",
    y = "Bookings",
    title = "Fair Play/ Tactical Fouls")+
  geom_text_repel(
    label= league.table$Club, 
    nudge_x = 0, nudge_y = -0.25, 
    aes(size = 3)
  )+
  scale_size_identity()
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "ycards">
### Yellow Cards
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(Club, YC))+
  geom_bar(stat = "identity",
           fill = league.table$Col)+
  theme_bw()+
  coord_flip()+
  labs(x = "Club",
       y = "Yellow Cards",
       title = "Premier League 2020-21 Yellow Cards")
```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

<p>&nbsp;</p>

<div id = "rcards">
### Red Cards
****************************

```{r,  message = FALSE}

ggplot(league.table, aes(Club, RC))+
  geom_bar(stat = "identity",
           fill = league.table$Col)+
  theme_bw()+
  coord_flip()+
  labs(x = "Club",
       y = "Red Cards",
       title = "Premier League 2020-21 Red Cards")

```
</div>

<div  style = "text-align: right;">
<a href = "#links">To Links</a>
</div>

</div>


By Rohan Pradhan (2018CSC1047)

<div  style = "text-align: right;">
<a href = "#top">To Top</a>
</div>

